---
title: "NYC Restraunt Insepection "
author: " Prisha Chowdhury 806755809"
date: "2025-11-25"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse) # I used AI to help familiarize myself with dplyr and its tools
library(scales)
```

# CORRELATION BETWEEN VIOLATION AND GRADE 

```{r}
# -----------------------------
# SUMMARY OF THE DATASET
# -----------------------------
df <- read.csv("newyork_restaurant_inspections.csv")
head(df)
str(df)
nrow(df)
ncol(df)
```
Hypothesis 1:
Worse grades are associated with higher number of violations (Correlation between grade and number of violations)

False Negative: 
Some A-graded restaurants have violations(especially critical ones)

```{r}
# -----------------------------
# KEEP ONLY NEEDED COLUMNS
# -----------------------------
# CAMIS                 -> Restaurant ID
# DBA                   -> Restaurant name
# INSPECTION.DATE       -> Date of inspection
# VIOLATION.CODE        -> Type of violation
# VIOLATION.DESCRIPTION -> Description of violation
# CRITICAL.FLAG         -> Whether violation was critical
# SCORE                 -> Numerical inspection score
# GRADE                 -> Assigned letter grade (A, B, C)
# ACTION                -> Action taken (e.g., closed, fined, etc.)

df_needed <- df[, c("CAMIS",
                   "DBA",
                   "INSPECTION.DATE",
                   "VIOLATION.CODE",
                   "VIOLATION.DESCRIPTION",
                   "CRITICAL.FLAG",
                   "SCORE",
                   "GRADE",
                   "ACTION")]
```

```{r}
# -----------------------------
# CLEAN DATA
# -----------------------------

# -------- CAMIS ---------
#Remove rows with missing ID
df_needed <- df_needed[!is.na(df_needed$CAMIS), ]

# -------- INSEPCTION.DATE ---------
#Convert to Date type and remove missing dates
df_needed$INSPECTION.DATE <- as.Date(df_needed$INSPECTION.DATE,
                                    format = "%m/%d/%Y")
#Check for missing values
sum(is.na(df_needed$INSPECTION.DATE))

# -------- VIOLATION.CODE ---------
#Check for missing values
sum(is.na(df_needed$VIOLATION.CODE))

# -------- CRITICAL.FLAG ---------
df_needed$CRITICAL.FLAG <- trimws(df_needed$CRITICAL.FLAG)   #remove spaces 
df_needed$CRITICAL.FLAG[df_needed$CRITICAL.FLAG == ""] <- NA  
df_needed$CRITICAL.FLAG <- toupper(df_needed$CRITICAL.FLAG)  #make everything uppercase for comparison

table(df_needed$CRITICAL.FLAG, useNA = "ifany")

# -------- GRADE ---------
#Remove row with missing Grade
df_needed <- df_needed[df_needed$GRADE != "", ]
table(df_needed$GRADE, useNA = "ifany")
```

```{r}
# -----------------------------
# DATA ANALYSIS
# -----------------------------

#Create a data frame to count number of violations
df_needed$INSPECTION_ID <- paste(df_needed$CAMIS,
                                 df_needed$INSPECTION.DATE,
                                 df_needed$GRADE,
                                 df_needed$SCORE,
                                 sep = "_")

inspection_df <- df_needed %>%
  group_by(INSPECTION_ID, CAMIS, INSPECTION.DATE, GRADE, SCORE) %>%
  summarize(
    num_violations = sum(!is.na(VIOLATION.CODE)),           
    num_critical   = sum(CRITICAL.FLAG == "CRITICAL", na.rm = TRUE)  # count critical violations
  ) %>%
  
  ungroup()
```

```{r}
# Calculate average number of violations per grade (Only A, B, C)
inspection_df <- inspection_df %>%
  filter(GRADE %in% c("A", "B", "C"))

inspection_df %>%
  group_by(GRADE) %>%
  summarize(
    mean_violations = mean(num_violations),
    n = n()
  )

# I used AI here to find out which hypothesis test is the best and how to use the function
# Correlation Test
inspection_df <- inspection_df %>%
  filter(GRADE %in% c("A", "B", "C")) %>%   
  mutate(
    grade_num = case_when(
      GRADE == "A" ~ 1,
      GRADE == "B" ~ 2,
      GRADE == "C" ~ 3,
      TRUE ~ NA_real_
    )
  )

cor.test(inspection_df$grade_num, inspection_df$num_violations)   #AI
```
Correlation coefficient: r = 0.6677886 
There is a moderately strong positive linear relationship between Grade and Number of Violations.
As grade worsens (A->B->C :: 1->2->3), the number of violation increases

p-value < 2.2e-16 which is extremely small. 
Null Hypothesis is rejected. (No linear relationship between Grade and Number of violations)

95% confidence interval: [0.6629520 0.6725695]
The true correlation in the population is between 0.663 and 0.73 (far from 0).
The relationship is strong and real

The data strongly supports H1: Worse grades are associated with higher number of violations

```{r}
# -----------------------------
# PLOT
# -----------------------------
# ggplot 2
inspection_df %>%
  filter(GRADE %in% c("A", "B", "C")) %>%
  ggplot(aes(x = GRADE, y = num_violations, fill = GRADE)) +
  geom_boxplot() +
  scale_fill_manual(values = c("skyblue", "gold", "salmon")) +
  labs(
    title = "Number of Violations by Grade",
    x = "Grade",
    y = "Number of Violations"
  ) +
  theme_minimal()
ggsave("num_violation_by_grade.png", width = 8, height = 6, dpi = 300)

# Scatter plot to see the trend
plot(inspection_df$grade_num, inspection_df$num_violations,
     xlab = "Grade (A=1, B=2, C=3)",
     ylab = "Number of Violations",
     main = "Correlation Between Grade and Violations")

abline(lm(num_violations ~ grade_num, data = inspection_df), col="red")

ggsave("scatter_corr_grade_violation.png", width=8, height=6, dpi=300)

```

```{r}
# -----------------------------
# FALSE NEGATIVES ANALYSIS
# -----------------------------
# Total number of A-grade inspections
total_A <- sum(inspection_df$GRADE == "A")

# A-graded inspections with ANY violations
A_any <- sum(inspection_df$GRADE == "A" & inspection_df$num_violations > 0)

# A-graded inspections with CRITICAL violations
A_crit <- sum(inspection_df$GRADE == "A" & inspection_df$num_critical > 0)

# Proportions
prop_any  <- A_any  / total_A
prop_crit <- A_crit / total_A

total_A
A_any
A_crit
prop_any
prop_crit
```
All inspections on A-graded restaurants had at least one documented violations.
81.36% of these inspections had at least one critical violation, meaning a violation that poses a direct risk to public health (e.g., improper food temperatures, evidence of rodents, or contamination hazards).

A grade suggests high food safety, yet the inspection documents violations, including critical ones. 
Consequently, an A grade does not guarantee the absence of health hazards; rather, it only reflects that the total violation score remained below the numerical threshold for a B or C grade.

# PERSISTENCE AND RECIDIVISM

## **Loading and Cleaning**

```{r}
restaurant <- read.csv("newyork_restaurant_inspections.csv")

# Remove columns: Latitude, Longitude, Community.Board. Council.District, Census.Tract, BIN, BBL, NTA, Location, BUILDING, STREET, ZIPCODE, PHONE, RECORD.DATE
restaurant <- restaurant[ , !(names(restaurant) %in% c("Latitude", "Longitude", "Community.Board", "Council.District", "Census.Tract", "BIN", "BBL", "NTA", "Location", "BUILDING", "STREET", "ZIPCODE", "PHONE","RECORD.DATE"))]

# Date formatting and filter all rows so that all inspections happened after 1996
restaurant$INSPECTION.DATE <- as.Date(restaurant$INSPECTION.DATE, format = "%m/%d/%Y")
restaurant <- restaurant %>% filter(INSPECTION.DATE >= as.Date("1996-01-01"))

# Remove all NAs for score
restaurant <- restaurant %>% filter(!is.na(SCORE))
```

```{r}
# We filter inspections by date. If different scores in the same inspection date, we will take the highest as the score
restaurant_unique <- restaurant %>%
  group_by(CAMIS, DBA, INSPECTION.DATE) %>%
  summarise(
    SCORE = max(SCORE),
    CRITICAL_VIOLATIONS = sum(CRITICAL.FLAG == "Critical", na.rm = TRUE),
    .groups = 'drop'
  ) %>%
  mutate(
    # Assign letter grade for each inspection
    GRADE = case_when(
      SCORE <= 13 ~ "A",
      SCORE <= 27 ~ "B",
      TRUE ~ "C"
    )
  ) %>%
  arrange(CAMIS, INSPECTION.DATE)

head(restaurant_unique)
```
##  **Analyzing the Data**

```{r}
# For each restaurant, compare current vs previous inspection
restaurant_persistence <- restaurant_unique %>%
  group_by(CAMIS) %>%
  arrange(INSPECTION.DATE) %>%
  mutate(
    score_prev = lag(SCORE),   # Previous inspection score
    score_change = SCORE - score_prev, 
    critical_prev = lag(CRITICAL_VIOLATIONS),
    
    # Indicates if critical violation is recurring
    critical_recur = ifelse(CRITICAL_VIOLATIONS > 0 & critical_prev > 0, 1, 0)
  ) %>%
  filter(!is.na(score_prev)) %>% #Remove first inspections
  ungroup()

head(restaurant_persistence)
```


```{r}
# Categorizes each inspection pair on whether the restaurant 'Improved', 'Worsened', or 'No Change' based on change in score 
score_summary <- restaurant_persistence %>%
  mutate(
    score_trend = case_when(
    score_change < 0 ~ "Improved",
    score_change > 0 ~ "Worsened",
    TRUE ~ "No Change"
    )
  ) %>%
  count(score_trend) %>%
  mutate(prop = n / sum(n)) # Proportion of each category

print(score_summary)
```

```{r}
## Computing proportions and count for recurrence of critical violations
critical_summary <- restaurant_persistence %>%
  summarise(
    total_pairs = n(),
    critical_recur_count = sum(critical_recur),
    prop_critical_recur = critical_recur_count / total_pairs
  )

print(critical_summary)
```
## Data Visualizations

```{r}
# Plot Score Trends (Improved / No Change / Worsened)
ggplot(score_summary, aes(x=score_trend, y=prop, fill=score_trend)) +
  geom_bar(stat = "identity") +
  scale_y_continuous(labels = percent_format()) +
  labs(title = "Improvement vs. Decline Across The Boroughs",
       x = "Score Trend (Compared to Previous Inspection)",
       y = "Proportion of Inspections") +
  theme_minimal()

ggsave("score_trend_barplot.png", width = 6, height = 4, dpi = 300)
```

```{r}
restaurant_yearly <- restaurant_persistence %>%
  mutate(year = format(INSPECTION.DATE, "%Y")) %>%
  group_by(year) %>%
  filter(year != "2015") %>%
  summarise(
    total_pairs = n(),
    critical_recur_count = sum(critical_recur),
    prop_critical_recur = critical_recur_count / total_pairs
  )


# Yearly Inspeciton Counts + Recurrence Rate on Log Scale
ggplot(restaurant_yearly, aes(x=year)) +
  geom_bar(
    aes(y=total_pairs), 
    stat="identity", 
    fill="lightblue", 
    alpha=0.7
  ) +
  geom_line(
    aes(y=prop_critical_recur * 1000, color="Recurrence Proportion", group=1),
    size=1.2
  ) +
  geom_point(
    aes(y=prop_critical_recur * 1000, color="Recurrence Proportion"), 
    size=2
  ) +
  scale_y_continuous(
    name="Number of Inspection Pairs",
    trans="log10",
    sec.axis = sec_axis(
      ~ . / 1000, 
      labels=scales::percent_format(), 
      name="Recurrence Proportion")
  ) +
  scale_color_manual(values= c("Recurrence Proportion"="darkred")) +
  labs(
    title="Log-Scaled Inspections Over Time", 
    x="Year", 
    color="Metric") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)
  )

ggsave("yearly_recidivism_plot.png", width =8, height=5, dpi = 300)
```
```{r}
# McNemar's Test for Recurrence of Critical Violations (AI-SUGGESTED IDEA)

# Contingency Table:
# Did a restaurant have critical violations previously vs currently?
tab <- table(
  prev = restaurant_persistence$critical_prev > 0,
  curr = restaurant_persistence$CRITICAL_VIOLATIONS >0
)

mcnemar.test(tab)
```
The McNemar's test gave a chi-squared statistic of 152.57 with 1 degree of freedom, as well as a p-value < 2.2e^(-16), which is far below any conventional significance level. These results indicate that restaurants with prior critical violations are significantly more likely to have violations in subsequent inspections, suggesting persistence rather than random occurrence.

## **Visualizing McNemar Contingency Table**

```{r}
#Converting the dataframe into a heatmap
mcnemar_df <- restaurant_persistence %>%
  mutate(
    prev = ifelse(critical_prev > 0, "Previous: Critical", "Previous: None"),
    curr = ifelse(CRITICAL_VIOLATIONS > 0, "Current: Critical", "Current: None")
  ) %>%
  count(prev, curr)

ggplot(mcnemar_df, aes(x = prev, y = curr, fill = n)) +
  geom_tile(color = "white") +
  geom_text(aes(label = n), size = 6, color = "black") +
  scale_fill_gradient(low = "#e0ecf4", high = "#0868ac") +
  labs(
    title = "Recurrence of Critical Violations",
    x = "",
    y = "",
    fill = "Count",
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "right",
    axis.text.x = element_text(size=12),
    axis.text.y = element_text(size=12)
  )

ggsave("critical_violation_heatmap.png", width=6, height=5, dpi=300)
```

# HIGH-RISK CUISINES AND BOROUGHS

```{r}
restaurant <- read.csv("newyork_restaurant_inspections.csv")
head(restaurant)
```

#cleaning for EDA 
```{r}
restaurant <-  restaurant[, c("CAMIS","BORO", "CUISINE.DESCRIPTION", "INSPECTION.DATE", "SCORE","VIOLATION.CODE","VIOLATION.DESCRIPTION","GRADE")]
restaurant$INSPECTION.DATE <- as.Date(restaurant$INSPECTION.DATE, format = "%m/%d/%Y")
restaurant$YEAR <- as.numeric(format(restaurant$INSPECTION.DATE, format = "%Y"))
restaurant <- restaurant[!is.na(restaurant$SCORE), ]
restaurant <- restaurant[restaurant$BORO != "" & restaurant$BORO != "0", ]
restaurant$CUISINE.DESCRIPTION <-
toupper(restaurant$CUISINE.DESCRIPTION) 
restaurant <- restaurant[restaurant$SCORE <= 60, ]
head(restaurant)
```
#Summaries 

##Most common cuisines
```{r}
cuisine_count <- sort(table(restaurant$CUISINE.DESCRIPTION), decreasing = TRUE)
top10_cuisine <-head(cuisine_count, 10)
```
American cuisine has the most occurence 

##Most inspected Boroughs 
```{r}
sort(table(restaurant$BORO), decreasing = TRUE)
```
Manhattan has the most occurance 
##Score Statistic 
```{r}
summary(restaurant$SCORE)
```

# Visualization 
```{r}
library(ggplot2)
```

## 1. Top 10 Most Common Cuisines
```{r}
barplot(top10_cuisine, main = "Top 10 Cuisine Types by Frequency", ylab = "Number of Inspections", color = "lightblue") 
```

## 2. Average Inspection Score by Borough
```{r}
avg_boro <- aggregate(SCORE ~ BORO, data = restaurant, FUN = mean)

ggplot(avg_boro, aes(x = reorder(BORO, SCORE), y = SCORE)) +
  geom_col(fill = "lightgreen") +
  labs(title = "Average Inspection Score by Borough", x = "Borough", y = "Average Score") +
  theme_minimal()

ggsave("avg_score_by_borough.png", width=8, height=6, dpi=300)

```

##3.Cuisines With the Highest Average Scores (Worst Overall Food Safety Performance)

```{r}
avg_cuisine <- aggregate(SCORE ~ CUISINE.DESCRIPTION, data = restaurant, mean)

worst10 <- avg_cuisine[order(avg_cuisine$SCORE, decreasing = TRUE),][1:10, ]

ggplot(worst10, aes(x = reorder(CUISINE.DESCRIPTION, SCORE), y = SCORE)) + 
geom_col(fill = "steelblue") + 
coord_flip() + 
labs(title = "Top 10 Highest-Risk Cuisines (Average Score)",
x = "Cuisine",
y = "Average Score") +
theme_minimal()

ggsave("high_risk_cuisines_avg.png", width=8, height=6, dpi=300)
```

##Top 10 Violation Codes
```{r}
viol_counts <- sort(table(restaurant$VIOLATION.CODE), decreasing = TRUE)
top10_viol <- head(viol_counts, 10)
top10_viol_df <- data.frame( ViolationCode = names(top10_viol), Count= as.numeric(top10_viol), stringsAsFactors = FALSE)

ggplot(top10_viol_df, aes(x = reorder(ViolationCode, Count), y = Count)) +
  geom_col(fill = "tomato") +
  coord_flip() +
  labs(
    title = "Top 10 Most Frequent Violation Codes",
    x = "Violation Code",
    y = "Count"
  ) +
  theme_minimal()

ggsave("frequent_violation_codes.png", width=8, height=6, dpi=300)
```
10F: Non-food contact surface or equipment made of unacceptable material, not kept clean, or not properly sealed, raised, spaced or movable to allow accessibility for cleaning on all sides, above and underneath the unit.
This is the most common violationn
## prediction 
```{r}
# Create BAD_GRADE variable: 1 = B/C, 0 = A
restaurant$BAD_GRADE <- ifelse(restaurant$GRADE %in% c("B", "C"), 1, 0)

# Keep relevant columns
ml_data <- restaurant[, c("BAD_GRADE", "SCORE")]

# Remove missing values
ml_data <- na.omit(ml_data)

# Fit simple logistic regression
model <- glm(BAD_GRADE ~ SCORE,
             data = ml_data,
             family = binomial)

summary(model)

```
The logistic regression shows that as inspection scores increase (indicating more violations), the probability of receiving a bad grade (B or C) increases significantly.

```{r}
plot(ml_data$SCORE, fitted(model),
     pch = 16, col = "#00000022",
     main = "Predicted Probability of Bad Grade vs. Score",
     xlab = "Inspection Score",
     ylab = "Predicted Probability")

```
**Observation** the logistic curve confirms a strong positive relationship between bad grades and violations: As inspection scores increase, the likelihood of receiving a bad grade increases sharply.

 
 
 
 
 
 
 
